# Impute.Rmd

This program was written in R Markdown written by Steve Simon. It requires
* R (no particular version)
---
title: "Special tricks with functions"
author: "Steve Simon"
date: "Thursday, April 28, 2016"
output: html_document
---

In R, there is a solid wall between what goes on in your R program
and what goes on inside the functions that you write for your
program. The functions are largely unaware of what is going on
in your program and the rest of your program is largely unaware
of what is going on in functions. There are ways, however, where
you can peek around the wall and pass information back and forth.

Let's first look at what functions can see in your main program.

```{r information_that_a_function_can_see}
vegas <- function() {
  if (lottery_winnings>0) {return("Congraluations!")}
     else return("Sorry!")
}
lottery_winnings <- 0
vegas()
lottery_winnings <- 1E6
vegas()
```

A function in R can see values that appear in your environment.

```{r informantion_that_your_main_program_cannot_see}
vegas <- function() {
  # You hit the jackpot.
  lottery_winnings <- 1e8
  print(paste("You've won", lottery_winnings,"dollars."))
  return("Congratulations!")
}
lottery_winnings <- 0
vegas()
print(lottery_winnings)
```

But your environment is not changed by changes inside your function.
In other words, what happens in Vegas, stays in Vegas.

```{r save-everything, timer=TRUE}
save.image("archive.RData")
```
