# Impute.Rmd

This program was written in R Markdown written by Steve Simon. It requires
* R (no particular version)
---
title: "Special tricks with functions"
author: "Steve Simon"
date: "Thursday, April 28, 2016"
output: html_document
---

In R, there is a solid wall between what goes on in your R program
and what goes on inside the functions that you write for your
program. The functions are largely unaware of what is going on
in your program and the rest of your program is largely unaware
of what is going on in functions. There are ways, however, where
you can peek around the wall and pass information back and forth.

Let's first look at what functions can see in your main program.

A function in R can see values that appear in your environment.

```{r information_that_your_function_can_see}
save.image("balckup.RData"); rm(list=ls()) # start with a blank slate.
vegas <- function() {
  if (lottery_winnings>0) {return("Congraluations!")}
     else return("Sorry!")
}
lottery_winnings <- 0
vegas()
lottery_winnings <- 1E6
vegas()
```

But any changes made inside a function disappear once the
function ends. In other words, what happens in Vegas stays
in Vegas. The only thing that the main program can see
is what the function returns.

```{r informantion_that_your_main_program_cannot_see}
rm(list=ls())
vegas <- function() {
  # You hit the jackpot.
  lottery_winnings <- 1e8
  print(paste("You've won", lottery_winnings,"dollars."))
  return("Congratulations!")
}
lottery_winnings <- 0
vegas()
print(lottery_winnings)
```

To allow changes inside a function to persist outside
the function, use the "<<-" assignment operator.

```{r information_that_your_main_program_can_see}
rm(list=ls())
vegas <- function() {
  # You hit the jackpot.
  lottery_winnings <<- 1e8
  print(paste("You've won", lottery_winnings,"dollars."))
  return("Congratulations!")
}
lottery_winnings <- 0
vegas()
print(lottery_winnings)
```

I like that a lot better, don't you?

Now there are problems in the opposite direction also. A function doesn't
know exactly what is being passed to it other than what the value itself
is.

```{r information_that_your_function cannot_see}
rm(list=ls())
vegas <- function(winnings) {
  print(paste("You have", winnings,"dollars in", "???"))
  return("Congratulations!")
}
lottery_winnings <- 1e6
vegas(lottery_winnings)
poker_winnings <- 1e7
vegas(poker_winnings)
```

```{r using_deparse_and_substitute}
rm(list=ls())
vegas <- function(winnings) {
  variable_name <- deparse(substitute(winnings))
  print(paste("You have", winnings,"dollars in", variable_name))
  return("Congratulations!")
}
lottery_winnings <- 1e6
vegas(lottery_winnings)
poker_winnings <- 1e7
vegas(poker_winnings)
```

These are artificial examples, but here is a real example where you
need to use the <<- assign operator and the deparse and substitute
functions.

Many of my programs accumulate a large number of intermediate
data.frames. I would like to store and document these data frames
so I could review things easily should things go wrong. The archive
function automates this process.

```{r archive_function}
rm(list=ls())
archive <- function(df) {
  # get the name of the data frame
  df_name <- deparse(substitute(df))
  print(df_name)
  if (!exists("arc")) arc <<- NULL
  arc[[df_name]] <<- df
  return(names(arc))
}
abc <- data.frame(x=LETTERS[1:3], y=letters[1:3])
archive(abc)
xyz <- data.frame(x=LETTERS[4:6], y=letters[4:6])
archive(xyz)
print(arc)
```

```{r save-everything, timer=TRUE}
save.image("archive.RData")
```
